---
title: "Ontario COVID-19"
output: 
  flexdashboard::flex_dashboard:
    vertical_layout: scroll
css: styles.css
runtime: shiny
---
```{r global, include=FALSE}
library(shiny)
library(shinydashboard)
library(plotly)
library(tidyverse)
library(leaflet)
library(lubridate)
library(gghighlight)
library(vroom)
library(sf)
# Load Data

ind_data <- vroom ("https://data.ontario.ca/dataset/f4112442-bdc8-45d2-be3c-12efae72fb27/resource/455fd63b-603d-4608-8216-7d8647f43350/download/conposcovidloc.csv",
                      col_types = cols(.default = "c")
                      ) %>% 
  mutate (Accurate_Episode_Date = ymd (Accurate_Episode_Date)) %>% 
  arrange (Reporting_PHU)

agg_data <- vroom ("https://data.ontario.ca/dataset/f4f86e54-872d-43f8-8a86-3892fd3cb5e6/resource/ed270bb8-340b-41f9-a7c6-e8ef587e6d11/download/covidtesting.csv")

agg_data  <-  agg_data %>% 
  mutate (`Reported Date` = ymd(`Reported Date`))

current_as_of <- tail (agg_data$`Reported Date`, 1)

# lubridate::ymd(current_as_of)
# lubridate::mdy(current_as_of)

```

About the Data User Group
================================================================

Sidebar {.sidebar data-width=220}
--------------------------------

We are a group of researchers and analysts who are interested in data science and would like to use our expertise to contribute to the understanding of COVID-19 in our communities.

#### Looking for data

One of the challenges we encountered trying to understand the spread of COVID-19 was finding a data source in a format that is easily accessible for analysis. When we were unable to locate such a file (and finding that the process to scrape data through R was too messy given the formats that the information has been released) we decided to take a manual approach. Using a few different sources, we have compiled data tables which are easily accessible in R (our favorite) and Python.

#### Opening the data from the source
A [googlesheet](https://docs.google.com/spreadsheets/d/1Y3_qYkTJK6Vnhw3RCOhOiwafm4-PQxd5l0Hxk4x1ZxE/edit#gid=1483529771) has been created and is being maintained using data from an Ontario government website and resources available on two Wikipedia pages. We will continue to update these tables until a more authoritative source of case records is made available, ideally by Public Health Ontario.

You can also access the data directly from the ministry's [webpage](https://data.ontario.ca/dataset/f4f86e54-872d-43f8-8a86-3892fd3cb5e6/resource/ed270bb8-340b-41f9-a7c6-e8ef587e6d11/download/covidtesting.csv) page.   

#### Resources: An invitation to explore and dive deeper

As we explore this data we will be sharing visualizations and insights on the Data User Group website. Our hope is that others will find our summaries useful. We extend an open invitation to others interested in data science to engage in additional analysis and use this data set for your own exploration

#### Data background and sources

You can read more about the project at the [Data User Group website](https://dugroup.ca/2020/03/19/covid19-understanding-ontario/).  

Column {data-width=600}
--------------------------------    

Data current as of `r current_as_of`

```{r Tiles}

fluidRow(
  
  column(2, align = "center",
         infoBox(title = HTML ("Cumulative Total Cases<br>"),
                 value = tail (agg_data$`Total Cases`, 1))),
  column(2, align = "center",
         infoBox(title = HTML ("Current Active Cases<br>"),
                 value = tail (agg_data$`Confirmed Positive`, 1))),
  column(2, align = "center",
         infoBox(title = HTML("Resolved Cases<br>"), 
                 value = tail (agg_data$Resolved, 1))),
  column(2, align = "center",
         infoBox(title = HTML("Number Tests<br>"), 
                 value = tail (agg_data$`Total patients approved for testing as of Reporting Date`, 1))),
  column(2, align = "center",
         infoBox(title = HTML("Cumulative Total Deaths<br>"), 
                 value = tail (agg_data$Deaths, 1)))
  )


```

<br>
<br>
<br>

```{r Status of Cases, fig.width=6}

# Set factor levels ----
factor_levels <- agg_data %>% 
  select("Reported Date", 
         "Cumulative Total Cases" = "Total Cases",
         "Current Active Cases" = "Confirmed Positive",
         "Resolved Cases" = "Resolved",
         "Cumulative Total Deaths" = "Deaths") %>% 
  tail (1) %>% 
  pivot_longer(-`Reported Date`,
               names_to = "group",
               values_to = "n") %>% 
  arrange (desc(n)) %>% 
  pull (group)

# aggregate data ----

long_data <- agg_data %>% 
  select("Reported Date", 
         "Cumulative Total Cases" = "Total Cases",
         "Current Active Cases" = "Confirmed Positive",
         "Resolved Cases" = "Resolved",
         "Cumulative Total Deaths" = "Deaths") %>% 
  pivot_longer(-`Reported Date`,
               names_to = "group",
               values_to = "n") %>% 
  mutate (group = parse_factor(group, 
                               levels = factor_levels)
          )

# Create chart -----
fluidRow(
  plot_ly(long_data, 
          x = ~`Reported Date`, 
          y = ~n,
          height = 400
  ) %>% 
    add_lines(color = ~group) %>% 
    layout(
      title = "Status of COVID-19 Cases in Ontario",
      xaxis = list (title = "Date"),
      yaxis = list (title = "Number of Cases",
                    range = c(0, max(agg_data$`Total Cases`, na.rm = TRUE) + 50)
      )
    )
)


```   

New Cases
================================================================

Sidebar {.sidebar data-width=200}
--------------------------------
One of the key indicators in determining when to ease restrictions in Ontario is the number of new cases. There needs to be a [consistent two–to-four week decrease in the number of new daily COVID‑19 cases](https://www.ontario.ca/page/framework-reopening-our-province#section-2). The chart below shows the total number of new cases since the beginning of February. 

Column {data-width=600}
--------------------------------

```{r New Cases, fig.width=5, fig.height=4}

# Key dates ----
today <- Sys.Date()
two_week_lag <- today - 14
rolling_mean <- tibbletime::rollify(mean, window = 7) # Function to calculate 7 dat average
shutdown <- ymd(as.Date("2021-04-03"))
phase_1 <- ymd(as.Date("2021-06-11"))
phase_2 <- ymd(as.Date("2021-06-30"))
phase_3 <- ymd(as.Date("2021-07-16"))

# Aggregate cases ----
new_cases <- agg_data %>% 
  arrange (`Reported Date`) %>% 
  dplyr::filter (`Reported Date` > "2020-02-28") %>% 
  mutate(`New Cases` = `Total Cases` - dplyr::lag(`Total Cases`),
         Time = ifelse (`Reported Date` > two_week_lag, "Previous 14 Days", paste0 ("Data up to ", two_week_lag))) %>%
  mutate (Time = parse_factor(Time),
          `7 Day Average` = rolling_mean(`New Cases`)) 

ymax <- max (new_cases$`New Cases`, na.rm = TRUE) + 50

seven_day_max <- round (max (new_cases$`7 Day Average`, na.rm = TRUE), 1)

# Plot ----
new_cases_plot <- new_cases %>% 
  ggplot(aes (x = `Reported Date`, y = `New Cases`)) + 
  geom_bar (aes(y = `New Cases`, fill = Time), stat = "identity") +
  geom_line (aes (y = `7 Day Average`, group = 1, color = "7 Day Average")) + 
  scale_fill_manual(name = "", values = c("grey60", "#0B6EA1")) +
  scale_color_manual(name = "", values = c("black")) +
  scale_x_date(expand = c(0,7),
               date_breaks = "1 month",
               date_labels =  "%b %d") +
  scale_y_continuous(limits = c(0, ymax)) + 
  # 7 Day Average ----
  geom_hline(yintercept = seven_day_max, size = 0.25) +
  labs (title = "Number of New COVID-19 Cases by Date",
        x = "Date") + 
  theme (panel.background = element_rect(fill = "NA"),
         legend.title = element_blank(),
         legend.position = "bottom"
         ) + 
  annotate("text", 
           size = 3, 
           x = ymd(as.Date("2020-06-15")), 
           y = seven_day_max + 225, 
           label =paste0 ("Highest 7 Day Average\n", seven_day_max)) + 
  # Shutdown ----
  geom_segment(aes(x = shutdown,
                   y = 0,
                   xend = shutdown,
                   yend = 3500),
               size = 0.25
               ) +
  annotate("text",
           size = 3,
           x = shutdown - 10,
           y = 4000,
           label = "Provincial\nShutdown") +
  # Phase 1 ----
  geom_segment(aes(x = phase_1,
                   y = 0,
                   xend = phase_1,
                   yend = 2000),
               size = 0.25
               ) +
  annotate("text",
           size = 3,
           x = phase_1,
           y = 2100,
           label = "Reopening\nPhase 1") +
  # Phase 2 ----
  geom_segment(aes(x = phase_2,
                   y = 0,
                   xend = phase_2,
                   yend = 1200),
               size = 0.25
               ) +
  annotate("text",
           size = 3,
           x = phase_2,
           y = 1300,
           label = "Reopening\nPhase 2") +
  # Phase 3 ----
  geom_segment(aes(x = phase_3,
                   y = 0,
                   xend = phase_3,
                   yend = 400),
               size = 0.25
               ) +
  annotate("text",
           size = 3,
           x = phase_3 - 1,
           y = 500,
           label = "Reopening\nPhase 3") +
  
  labs (title = "Number of New COVID-19 Cases by Date",
        x = "Date") + 
  theme (panel.background = element_rect(fill = "NA"),
         legend.title = element_blank()
         ) 
  

fluidPage(
  renderPlotly(ggplotly(new_cases_plot, 
                        width = 1000,
                        height = 600) %>%
                 layout(legend = list(orientation = "h", x = 0.3, y = -0.2))
))

```

Cases by Schools
================================================================

Sidebar {.sidebar data-width=220}
--------------------------------

```{r include=FALSE}
library(tidyverse)
library(leaflet)
library(leaflet.minicharts)
library(readxl)
# setwd("C:/Users/grousell/Dropbox/COVID-19/Ontario-COVID-19-Research/COVID-19/")
school_covid <- read_csv ("school_covid.csv",
                          col_types = cols(collected_date = col_character(),
                                           reported_date = col_character(),
                                           school_board = col_character(),
                                           school = col_character(),
                                           municipality = col_character(),
                                           confirmed_student_cases = col_double(),
                                           confirmed_staff_cases = col_double(),
                                           total_confirmed_cases = col_double(),
                                           confirmed_unspecified_cases = col_double(),
                                           lat = col_double(),
                                           lon = col_double()
                          ),
                          locale = readr::locale(encoding = "UTF-8")
) %>%
  rename (Students = confirmed_student_cases,
          Staff = confirmed_staff_cases,
          Unspecified = confirmed_unspecified_cases)


n_students <- sum (school_covid$Students, na.rm = TRUE)
n_staff <- sum (school_covid$Staff, na.rm = TRUE)
n_unspecified <- sum (school_covid$Unspecified, na.rm = TRUE)
n_total <- sum (school_covid$total_confirmed_cases, na.rm = TRUE)
n_schools <- length (school_covid$school)
n_school_boards <- length(unique(school_covid$school_board))


# group <- c("Total Cases in Schools", "Students", "Staff", "Unspecified", "Schools", "School Boards")
# n <- c(n_total, n_students, n_staff, n_unspecified, n_schools, n_school_boards)
# 
# test <- data.frame (group, n)

```

This map shows the school boards with reported cases of COVID-19 as reported to the Ministry of Health. The dataset can be found [here](https://data.ontario.ca/dataset/b1fef838-8784-4338-8ef9-ae7cfd405b41/resource/8b6d22e2-7065-4b0f-966f-02640be366f2/download/schoolsactivecovid.csv). The Data User Group has started and maintained a data sheet with the schools' geocode [here](https://raw.githubusercontent.com/DUGroup/Ontario-COVID-19-Research/master/Data/schools_geocode.csv).

The circle shows where confirmed cases have occurred with either a staff member or student. Clicking the circle will show the number of confirmed cases for both staff and students.  

This data was current as of April 3, 2021 when the government implemented a province-wide shutdown.  

Column {data-width=100}
--------------------------------


```{r}
infoBox(title = HTML ("Total Cases in Schools<br>"),
        value = n_total, 
        width = 8,
        icon = shiny::icon("chart-pie"))

infoBox(title = HTML ("Students<br>"),
                 value = n_students,
        icon = shiny::icon("child"))

infoBox(title = HTML ("Staff<br>"),
                 value = n_staff,
        icon = shiny::icon("chalkboard-teacher"))

infoBox(title = HTML ("Unspecified<br>"),
                 value = n_unspecified,
        icon = shiny::icon("question-circle"))

infoBox(title = HTML ("School Boards<br>"),
                 value = n_school_boards,
        icon = shiny::icon("school"))

infoBox(title = HTML ("Schools<br>"),
                 value = n_schools,
        icon = shiny::icon("chalkboard"))

```


Column {.tabset}
--------------------------------

### Map of Active School Cases
```{r include = TRUE, fig.width=8, fig.height=6}

output$map_school <- renderLeaflet({
  leaflet(school_covid) %>%
    addTiles() %>%
    # setView(lat = 46.5830386,
    #         lng = -81.639553,
    #         zoom = 6) %>% 
    addMinicharts(
      school_covid$lon, school_covid$lat,
      type = "pie",
      chartdata = school_covid[, c("Students", 
                                   "Staff",
                                   "Unspecified")],
      colorPalette = c("green", "purple", "blue"),
      showLabels = TRUE,
      layerId = paste0(school_covid$school, " - ", school_covid$school_board),
      labelText = paste0(school_covid$school, " - ", school_covid$school_board), 
      width = school_covid$total_confirmed_cases * 8
    )
}
)

leafletOutput("map_school", 
              height = "300") 


```

### List of Schools

```{r SchoolTable, include=TRUE}

DT::renderDataTable (
  DT::datatable(school_covid %>% select (`Report Date` = reported_date,
                                         `School Board` = school_board, 
                                         School = school,
                                         Students,
                                         Staff, 
                                         Unspecified,
                                         Total = total_confirmed_cases), 
                options = list(pageLength = 40, 
                               scrollX = TRUE,
                               scrollY = '400px',
                               autoWidth = FALSE),
                rownames = FALSE)
) 


```


### School Cases over Time

```{r include = TRUE, fig.width=8, fig.height=6}

school_covid_all <- read_csv ("https://data.ontario.ca/dataset/b1fef838-8784-4338-8ef9-ae7cfd405b41/resource/8b6d22e2-7065-4b0f-966f-02640be366f2/download/schoolsactivecovid.csv",
                              col_types = c (.default = "c"))


palette <- c("#C5E8DD", "#FDC3AC", "#8DA0CB", "#F6D3E8")

school_over_time <- school_covid_all %>% 
  mutate (reported_date = ymd (reported_date)) %>% 
  group_by (reported_date) %>%
  summarise (Student = sum (as.numeric(confirmed_student_cases)),
             Staff = sum (as.numeric(confirmed_staff_cases)), 
             Unspecified = sum (as.numeric (confirmed_unspecified_cases))
             ) %>% 
  pivot_longer(cols = c (Student, Staff, Unspecified)) %>% 
  rename (Date = reported_date, 
          Group = name, 
          N = value)

school_over_time_totals <- school_over_time %>% 
  group_by (Date) %>% 
  summarise (Total = sum (N, na.rm = TRUE)) 

school_over_time_plot <- school_over_time %>% 
  ggplot (aes (x = Date, y = N, fill = Group)) + 
  geom_col() + 
  scale_fill_manual(values = palette) + 
  geom_text (aes (label = N), 
             position = position_stack(vjust = 0.5), 
             size = 1.5
             ) + 
  geom_text (aes (x = Date, y = Total + 20, label = Total, fill = NULL),
             data = school_over_time_totals,
             size = 2
             ) +
  labs (title = "Active School Cases of COVID-19",
        subtitle = "Over Time",
        x = "Reported Date", 
        y = "Number of Cases") + 
  theme(legend.title=element_blank(),
        legend.position = "bottom",
        panel.background = element_rect(fill="NA")
        )
  
renderPlotly(ggplotly(school_over_time_plot, 
                      tooltip = c("label", "fill", "x")) %>% 
               layout(legend = list(orientation = "h", x = 0.3, y = -0.2))
)


```




Public Health Units
================================================================

Sidebar {.sidebar data-width=250}
--------------------------------
The data is based on the [provincial data set](https://data.ontario.ca/en/dataset/confirmed-positive-cases-of-covid-19-in-ontario) and reflects the last submission by the Public Health Unit.  

```{r Selector}

selectInput("phu", 
            "Public Public Health Unit",
            choices = sort(unique(ind_data$Reporting_PHU))
            )  

```

Column {data-width=900}
--------------------------------

```{r Total Cases PHU}
  
fluidRow(
  renderPlotly(
    ggplotly(
      # AGGREGATE DATA -------------  
      ind_data %>%
        group_by (Accurate_Episode_Date , Reporting_PHU) %>%
        count() %>%
        dplyr::filter (Reporting_PHU == input$phu ) %>%
        # dplyr::filter (Reporting_PHU == "Algoma Public Health Unit" ) %>%
        arrange (Accurate_Episode_Date) %>%
        ungroup() %>%
        mutate (total_n = cumsum(n),
                Accurate_Episode_Date = ymd (Accurate_Episode_Date)) %>% 
        # START PLOT ------------------------------------
        {
          ggplot (., aes (x = Accurate_Episode_Date, 
                          y = total_n, 
                          colour = Reporting_PHU)) + 
            geom_line(colour = "#0B6EA1") +
            scale_x_date(expand = c(0,7),
                         date_breaks = "2 weeks",
                         date_labels =  "%b %d") +
            labs (title = paste0 (input$phu, "\nTotal Number of COVID-19 Cases"),
                  x = "", 
                  y = "Number of Cases") + 
            theme (panel.background = element_rect(fill = "NA"),
                   legend.title = element_blank(),
                   plot.title = element_text(hjust = 0.5),
                   axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
                   )
        }
      # END GGPLOT ------------------------------------
      ) %>%
      layout(legend = list(orientation = "h", x = 0.3, y = -0.2),
             width = 900,
             height = 400
             )
    ) %>% 
      bindCache (input$phu)
  )


  

```


```{r New Cases by Region, include=TRUE}

fluidRow(
  renderPlotly(
    ggplotly(
      # AGGREGATE DATA -------------  
      ind_data %>%
        dplyr::filter (Accurate_Episode_Date >= "2020-01-01") %>%
        dplyr::filter(Reporting_PHU == input$phu) %>%
        group_by (Accurate_Episode_Date) %>%
        # dplyr::filter(Reporting_PHU == "Toronto Public Health") %>%
        count() %>%
        rename(Date = Accurate_Episode_Date, 
               `New Cases` = n) %>% 
        mutate (Date = as.Date(Date)) %>% 
        complete(Date = seq.Date(as.Date("2020-01-01"), max(Date), by="day")) %>% 
        replace_na(list(`New Cases` = 0)) %>% 
        distinct() %>% 
        ungroup() %>% 
        mutate (`7 Day Average` = rolling_mean(`New Cases`)) %>%
        mutate (Time = ifelse (Date > two_week_lag, 
                               "Previous 14 Days", 
                               paste0 ("Data up to ", two_week_lag))) %>% 
      # START PLOT ------------------------------------
      {
        ggplot (., aes (x = Date, y = `New Cases`)) +
          geom_bar (stat = "identity") + 
          geom_bar (aes(y = `New Cases`, fill = Time), stat = "identity") +
          geom_line (aes (y = `7 Day Average`, group = 1, color = "7 Day Average")) + 
          scale_fill_manual(name = "", values = c("grey60", "#0B6EA1")) +
          scale_color_manual(name = "", values = c("black")) +
          scale_x_date(expand = c(0,7),
                       date_breaks = "2 weeks",
                       date_labels =  "%b %d") +
          scale_y_continuous(limits = c(0, max(.$`New Cases`))) +
          # Shutdown -----------------------------------
        geom_segment(aes(x = shutdown,
                         y = 0,
                         xend = shutdown,
                         yend = max(.$`New Cases`) *0.9)
                     ) +
          annotate("text",
                   size = 3,
                   x = shutdown - 10,
                   y = max(.$`New Cases` * 0.95),
                   label = "Provincial\nShutdown") +
          # Phase 1 -----------------------------------
        geom_segment(aes(x = phase_1,
                         y = 0,
                         xend = phase_1,
                         yend = max(.$`New Cases`) *0.75)
                     ) +
          annotate("text",
                   size = 3,
                   x = phase_1,
                   y = max(.$`New Cases` * 0.80),
                   label = "Reopening Phase 1") +
          # Phase 2 -----------------------------------
        geom_segment(aes(x = phase_2,
                         y = 0,
                         xend = phase_2,
                         yend = max(.$`New Cases` * 0.65))
                     ) +
          annotate("text",
                   size = 3,
                   x = phase_2,
                   y = max(.$`New Cases` *0.70),
                   label = "Reopening Phase 2") +
          # Phase 3 -----------------------------------
        geom_segment(aes(x = phase_3,
                         y = 0,
                         xend = phase_3,
                         yend = max(.$`New Cases` * 0.55))
                     ) +
          annotate("text",
                   size = 3,
                   x = phase_3,
                   y = max(.$`New Cases` * 0.6),
                   label = "Reopening Phase 3") +
          # Labels --------------------------
        labs (title = paste0 (input$phu, "\nNew COVID-19 Cases by Date"),
              x = "", 
              y = "Number of New Cases") + 
          # Theme --------------------------
        theme (panel.background = element_rect(fill = "NA"),
               legend.title = element_blank(),
               plot.title = element_text(hjust = 0.5),
               axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
        )
        }
    # END GGPLOT ------------------------------------
    ) %>%
      layout(legend = list(orientation = "h", x = 0.3, y = -0.2),
             width = 900,
             height = 400
             )
    ) %>% 
    bindCache (input$phu)
  )


```


Map of Cases by PHU
================================================================

Sidebar {.sidebar data-width=220}
--------------------------------

The map provides data only for those cases that have been submitted to the Ministry of Health. The full dataset can be found [here](https://data.ontario.ca/en/dataset/confirmed-positive-cases-of-covid-19-in-ontario). 


```{r include=FALSE}
source ("maps.R")
```


Column {data-width=600}
--------------------------------

```{r fig.width=9, fig.height=6}

map_phu

```

