---
title: "Ontario COVID-19 Tracking"
output: flexdashboard::flex_dashboard
css: styles.css
runtime: shiny
---
```{r global, include=FALSE}
library(shiny)
library(shinydashboard)
library(plotly)
library(tidyverse)
library(leaflet)
library(lubridate)

df <- read_csv ("https://raw.githubusercontent.com/DUGroup/Ontario-COVID-19-Research/master/Data/ontario_corona_cases.csv")

dates_all <- read_csv("https://raw.githubusercontent.com/DUGroup/Ontario-COVID-19-Research/master/Data/dates_all.csv") 

prov_agg <- read_csv ("https://raw.githubusercontent.com/DUGroup/Ontario-COVID-19-Research/master/Data/provincial_aggregrate.csv")

phu <- read_csv("https://raw.githubusercontent.com/DUGroup/Ontario-COVID-19-Research/master/Data/phu_lat_long.csv")

# Clean data --------------------------------------------------------------

df <- df %>% 
    mutate (Public_Health_Unit = ifelse (str_detect(Public_Health_Unit, "RichmondHill"), "York",
                                         ifelse (str_detect(Public_Health_Unit, "Newmarket"), "York",
                                                 Public_Health_Unit))) %>% 
    mutate (Public_Health_Unit = ifelse(str_detect (Public_Health_Unit, "Haliburton"),
                                        "Haliburton,Kawartha,Pine Ridge",
                                        Public_Health_Unit)) %>% 
  mutate (Public_Health_Unit = ifelse(str_detect (Public_Health_Unit, "Kingston"),
                                        "Kingston, Frontenac and Lennox & Addington",
                                        Public_Health_Unit)) %>% 
  mutate (Public_Health_Unit = ifelse(str_detect (Public_Health_Unit, "Porcupine"),
                                        "Porcupine District",
                                      Public_Health_Unit)) %>% 
  mutate (Public_Health_Unit = ifelse(str_detect (Public_Health_Unit, "Sudbury"),
                                        "Sudbury and District",
                                      Public_Health_Unit)) %>%
  mutate(Public_Health_Unit = ifelse(Public_Health_Unit == "pending" |
                                           Public_Health_Unit == "NotAvailable", "Pending or Not Available", Public_Health_Unit))

# df %>% group_by (Public_Health_Unit) %>%  count () %>% print (n = Inf)

df_province <- df %>% 
    group_by (Date) %>% 
    count () %>% 
    ungroup() %>% 
    mutate (cumsum = cumsum(n),
            region = "Province") 

# Create master data frame
cases_by_region <- dates_all %>%
    left_join (df, by = c ("Date", "Public_Health_Unit")) %>%
    group_by(Public_Health_Unit) %>%
    mutate(Cases = cumsum (!is.na(Case_number))) %>%
    ungroup() %>%
    mutate (Public_Health_Unit = parse_factor(Public_Health_Unit))

# # Create linking colum for map
# 
# cases_by_region <- cases_by_region %>% 
#   mutate (new_phu = dplyr::recode(as.character (Public_Health_Unit), 
#                            "Middlesex London" = "Middlesex"))

cases_by_region_map <- read_csv ("https://raw.githubusercontent.com/DUGroup/Ontario-COVID-19-Research/master/Data/regional_data.csv"
                ) %>% 
  mutate (Date = ymd(Date)) %>% 
  # remove Simoce as it's now combined with Muskoks
  dplyr::filter (Region != "Simcoe") %>% 
  dplyr::filter(Region != "Muskoka") %>% 
  # filter for most recent entry by Region
  group_by(Region) %>% top_n(1, Date)

n_pending <- df %>% 
  dplyr::filter(Public_Health_Unit == "Pending or Not Available") %>% 
  count ()


```


    
Column {data-width=200}
-------------------------------------
    
### 

#### Ontario Data User Group (DUG)

We are a group of researchers and analysts who are interested in data science and would like to use our expertise to contribute to the understanding of COVID-19 in our communities.

#### Looking for data

One of the challenges we encountered trying to understand the spread of COVID-19 was finding a data source in a format that is easily accessible for analysis. When we were unable to locate such a file (and finding that the process to scrape data through R was too messy given the formats that the information has been released) we decided to take a manual approach. Using a few different sources, we have compiled data tables which are easily accessible in R (our favorite) and Python.

#### Opening the data from the source
A [googlesheet](https://docs.google.com/spreadsheets/d/1Y3_qYkTJK6Vnhw3RCOhOiwafm4-PQxd5l0Hxk4x1ZxE/edit#gid=1483529771) has been created and is being maintained using data from an Ontario government website and resources available on two Wikipedia pages. We will continue to update these tables until a more authoritative source of case records is made available, ideally by Public Health Ontario.

You can also access the data directly from DUG's [github](https://raw.githubusercontent.com/DUGroup/Ontario-COVID-19-Research/master/Data/ontario_corona_cases.csv) page.   

#### Resources: An invitation to explore and dive deeper

As we explore this data we will be sharing visualizations and insights on the Data User Group website. Our hope is that others will find our summaries useful. We extend an open invitation to others interested in data science to engage in additional analysis and use this data set for your own exploration

#### Data background and sources

You can read more about the project at the [Data User Group website](https://dugroup.ca/2020/03/19/covid19-understanding-ontario/).  


Column {.tabset} 
-------------------------------------

### Ontario

Data current as of `r Sys.Date()`

```{r}

fluidRow(
  column(3, align = "center",
         infoBox(title = h4("Confirmed Positive"),
                 value = h4(length(df$Case_number)))),
  column(3, align = "center",
         infoBox(title = h4("Resolved"), 
                 value = h4(prov_agg[4,2]))),
  column(3, align = "center",
         infoBox(title = h4("Number Tests"), 
                 value = h4(prov_agg[6,2]))),
  column(3, align = "center",
         infoBox(title = h4("Deceased"), 
                 value = h4(prov_agg[5,2])))
  )


```
<br>
<br>
<br>

```{r}
plot_ly(df_province, 
        x = ~Date, 
        y = ~cumsum,
        height = 400) %>% 
  add_lines() %>% 
  layout(
    title = "Total Number of COVID-19 Cases in Ontario",
    xaxis = list (title = "Date"),
    yaxis = list (title = "Number of Cases",
                  range = c(0, max(df_province$cumsum) + 50)
                  )
    )
    

```   
 

### Cases by Public Health Unit

The dataset currently incudes `r n_pending` cases listed as Pending or Not Available. The data below represents only those cases that have been identified by Public Health Unit. 

```{r}

phu_list <- sort (unique(df$Public_Health_Unit))

selectInput("phu", 
            "Public Heatlh Unit",
            choices = phu_list)

```


    
```{r}

renderPlotly(
plot_ly(cases_by_region %>% dplyr::filter (Public_Health_Unit == input$phu), 
        x = ~Date, 
        y = ~Cases,
        height = 400) %>% 
            add_lines (color = ~Public_Health_Unit) %>% 
            layout(
                title = "Total Number of COVID-19 Cases by Public Health Unit",
                xaxis = list (title = "Date"),
                yaxis = list (title = "Number of Cases")
            )
)
    
```

### Map of Cases by Public Health Unit

The map provides data only for those cases that have an identified health unit. This data has been collected through Wikipedia and various news reports. The full dataset can be found at the [DUG googlesheets](https://docs.google.com/spreadsheets/d/1Y3_qYkTJK6Vnhw3RCOhOiwafm4-PQxd5l0Hxk4x1ZxE/edit) or [github](https://github.com/DUGroup/Ontario-COVID-19-Research/blob/master/Data/regional_data.csv) page.  

The purple circles show the number of confirmed COVID-19 cases in that region. The larger the circle, the more cases. Cicking on the circle will show the name of the Public Health Unit and the number of cases. 

```{r fig.height=5}

# total_cases_region <- df %>% 
#   group_by (Public_Health_Unit) %>% 
#   count ()
# 
# phu <- phu %>% 
#   left_join(total_cases_region)

leaflet(data = cases_by_region_map) %>%
  addTiles() %>%  
  addCircleMarkers(~long, ~lat, 
                   radius = ~sqrt(Cases) + 3,
                   label = ~as.character(paste0(Region, " ", Cases, " cases")),
                   popup = ~as.character(paste0(Region, " ", Cases, " cases")),
                   stroke = FALSE, 
                   fillOpacity = 0.7,
                   color = "purple")

```

